\include{commons}
\lecturetitle{Lecture 9a: Spans and Vector Spaces} 

\begin{frame}
  \frametitle{Review: Linear combinations}

  \begin{block}{Definition}
    For any scalars
    \[
    \alpha_1,\alpha_2,\ldots,\alpha_m
    \]
    and vectors
    \[
    \uv_1,\uv_2,\ldots,\uv_m,
    \]
    we say that
    \[
    \alpha_1\uv_1 + \alpha_2\uv_2 + \cdots + \alpha_m \uv_m
    \]
    is a \textcolor{red}{\bf linear combination} of $\uv_1,\ldots,\uv_m$.
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Review: Span}

  \begin{block}{Definition}
    A set of all linear combination of vectors $\uv_1,\uv_2,\ldots,\uv_m$ is called the \textcolor{red}{\bf span} of that set of vectors.

    It is denoted by $\mathrm{Span} \{\uv_1,\uv_2,\ldots,\uv_m\}$.
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Example 1}
  Is $\mathrm{Span}\ \{[1,2],[2,5]\} = \rf^2$?
  \vspace{2.5in}
\end{frame}

\begin{frame}
  \frametitle{Example 2}
  Is $\mathrm{Span}\ \{[1,0,1],[1,1,0],[2,3,4]\} = \rf^3$?
  \vspace{2.5in}
\end{frame}

\begin{frame}
  \frametitle{Example 3}
  Is $\mathrm{Span}\ \{[1,0,1],[1,1,0],[4,2,2]\} = \rf^3$?
  \vspace{2.5in}
\end{frame}

\begin{frame}
  \frametitle{Elements in a vector}
  \begin{itemize}
  \item We see examples of vectors over $\rf$.
  \item However, elements in a vector can be from other sets with
    appropriate property.  (I.e., they should behave a real numbers.)
  \item What do we want from an element in a vector?
    \begin{itemize}
    \item We should be able to perform addition, subtraction, multiplication, and division.
    \item Operations should be commutative and associative.
    \item Additive and multiplicative identity should exist.
    \item Addition and multiplication should have inverses.
    \end{itemize}

    \item We refer to a set with these properties as a {\bf field}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{A field}

  \begin{block}{Definition}
    A set $\ff$ with two operations $+$ and $\times$ (or $\cdot$) is a
    \textcolor{red}{\bf field} iff these operations satisfy the
    following properties:
    \begin{itemize}
    \item (Associativity): $(a+b)+c = a+(b+c)$ and $(a\cdot b)\cdot c = a\cdot(b\cdot c)$
    \item (Commutativity): $a+b=b+a$ and $a\cdot b=b\cdot a$
    \item (Identities): There exist two elements $0\in\ff$ and $1\in\ff$ such that $a+0 = a$ and $a\cdot 1 = a$
    \item (Additive inverse): For every element $a\in \ff$, there is an element $-a\in \ff$ such that $a+(-a) = 0$
    \item (Multiplicative inverse): For every element $a\in \ff\setminus\{0\}$, there is an alement $a^{-1}$ such that $a\cdot a^{-1}=1$
    \item (Distributive): $a\cdot(b+c)=a\cdot b + a\cdot c$
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Another useful field: $GF(2)$}
  $GF(2) = \{0,1\}$.  I.e., it is a ``bit'' field.

  What are $+$ and $\cdot$ in $GF(2)$?

  \pause

  \begin{itemize}
  \item We define $b_1+b_2$ to be XOR.
    \[
    \begin{array}{c}
      0 + 0 = 0 \\ 
      0 + 1 = 1 + 0 = 1 \\ 
      1 + 1 = 0
    \end{array}
    \] \pause
  \item We define $b_1\cdot b_2$ to be standard multiplication.
    \[
    \begin{array}{c}
      0 \cdot 0 = 0\cdot 1 = 1\cdot 0 = 0 \\
      1\cdot 1 = 1
    \end{array}
    \] 
    
  \end{itemize}

  You can check that $GF(2)$ satisfies the axioms of fields.
\end{frame}

\begin{frame}
  \frametitle{$2\times 2$ Lights out}
\end{frame}

\begin{frame}
  \frametitle{Can you solve $2\times 2$ Lights out?}

  Let
  $\uv_1=[1,1,1,0]$, 
  $\uv_2=[1,1,0,1]$, 
  $\uv_3=[1,0,1,1]$, and 
  $\uv_4=[0,1,1,1]$.

  \vspace{0.2in}
  
  Given $\vect{b}=[b_1,b_2,b_3,b_4]$, can you always find
  $a_1,a_2,a_3,a_4\in GF(2)$ such that
  \[
  a_1\cdot \uv_1 + 
  a_2\cdot \uv_2 + 
  a_3\cdot \uv_3 + 
  a_4\cdot \uv_4 = \vect{b} ?
  \]

  \pause

  \vspace{0.2in}

  {\bf Same question:} Is $\mathrm{Span}\;\{\uv_1,\uv_2,\uv_3,\uv_4\} = GF(2)^4$?
\end{frame}


\begin{frame}
  \frametitle{Can you solve $2\times 2$ Lights out?}

  Let's try with an example.  Let $\vect{b}=[1,0,0,0]$.  Can you find
  $a_1,a_2,a_3,a_4\in GF(2)$ such that
  \[
  a_1\cdot \uv_1 + 
  a_2\cdot \uv_2 + 
  a_3\cdot \uv_3 + 
  a_4\cdot \uv_4 = \vect{b} ?
  \]

  \vspace{2in}

\end{frame}


\begin{frame}
  \frametitle{Can you solve $2\times 2$ Lights out?}

  Since
  \[
  [1,0,0,0], [0,1,0,0], [0,0,1,0], [0,0,0,1]\in {\mathrm{Span}}\ \{\uv_1,\uv_2,\uv_3,\uv_4\},
  \]
  and
  \pause
  \[
  \mathrm{Span}\ \{[1,0,0,0], [0,1,0,0], [0,0,1,0], [0,0,0,1]\} = GF(2)^4,
  \]
  \pause
  what can we say about ${\mathrm{Span}}\ \{\uv_1,\uv_2,\uv_3,\uv_4\}$?

\end{frame}


\begin{frame}
  \frametitle{Generators}

  \begin{block}{Definition}
    Let $\V$ be a set of vectors.  Consider vectors $\uv_1,\uv_2,\ldots,\uv_n$.

    If $\mathrm{Span}\ \{\uv_1,\uv_2,\ldots,\uv_n\} = \V$, we say that
    \begin{itemize}
    \item $\{\uv_1,\uv_2,\ldots,\uv_n\}$ is a {\bf generating set} for $\V$
    \item vectors $\uv_1,\uv_2\ldots,\uv_n$ are {\bf generators} for $\V$
    \end{itemize}
  \end{block}

  \pause

  {\bf Examples}
  \vspace{1.5in}
\end{frame}

\begin{frame}
  \frametitle{Standard generators}

  Note that $\{[1,0,0,0], [0,1,0,0], [0,0,1,0], [0,0,0,1]\}$ are
  generators for $GF(2)^4$.  Why?

  \vspace{0.5in}
  \pause

  They are called {\bf standard generators} for $GF(2)^4$, written as $\vect{e}_1,\vect{e}_2,\vect{e}_3,\vect{e}_4$.
  \pause

  \vspace{0.2in}

  For $\rf^n$, we also have
  $[1,0,0,\ldots,0],[0,1,0,\ldots,0],[0,0,1,\ldots,0],\ldots,[0,0,0,\ldots,1]$
  as standard generators.
  
\end{frame}

\begin{frame}
  \frametitle{Generators and spans}
  \begin{lemma}
    Consider vectors $\uv_1,\uv_2,\ldots,\uv_n$.
    If $\vv_1,\vv_2,\ldots,\vv_k$ are generators for $\V$, and for each $i$,
    \[
    \vv_i\in \mathrm{Span}\ \{\uv_1,\uv_2,\ldots,\uv_n\},
    \]
    we have that $\V \subseteq \mathrm{Span}\ \{\uv_1,\uv_2,\ldots,\uv_n\}$.
  \end{lemma}

  \vspace{2in}
\end{frame}

\begin{frame}
  \frametitle{Adding a vector into a span}
  \begin{lemma}
    Consider vectors $\uv_1,\uv_2,\ldots,\uv_n$.
    If $\vv\in \mathrm{Span}\ \{\uv_1,\uv_2,\ldots,\uv_n\}$, then
    \[
    \mathrm{Span}\ \{\uv_1,\uv_2,\ldots,\uv_n,\vv\}
    =
    \mathrm{Span}\ \{\uv_1,\uv_2,\ldots,\uv_n\}
    \]
  \end{lemma}

  \vspace{2in}
\end{frame}

\begin{frame}
  \frametitle{Geometry of spans: in $\rf^2$}
\end{frame}

\begin{frame}
  \frametitle{Geometry of spans: in $\rf^3$}
\end{frame}

\begin{frame}
  \frametitle{Two representations}

  There are two ways to represent a line, a plane, and a (hyper)plane,
  passing through the origin:
  \begin{itemize}
  \item as a span of vectors
  \item as solutions of a system of homogeneous linear equations.
  \end{itemize}

  \pause
  \vspace{0.2in}

  What are common properties of these geometric objects?
  \pause
  \begin{itemize}
  \item they pass through the origin,
  \item if vector $\uv$ is in the objects, $\alpha\uv$ for any scalar $\alpha$ is also in the objects, and
  \item if $\uv$ and $\vv$ are in the objects, $\uv+\vv$ is also in
    the objects.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Vector spaces}
  \begin{block}{Definition}
    A set $\V$ of vectors over $\ff$ is a \textcolor{red}{\bf vector space} iff
    \begin{itemize}
    \item \textcolor{blue}{(V1)} $\vect{0}\in\V$,
    \item \textcolor{blue}{(V2)} for any $\uv\in\V$,
      \[
      \alpha\cdot\uv\in\V
      \]
      for any
      $\alpha\in\ff$, and
    \item \textcolor{blue}{(V3)} for any $\uv,\vv\in\V$,
      \[
      \uv+\vv\in\V.
      \]
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Span of vectors is a vector space}
  Consider $n$-vectors $\uv_1,\uv_2,\ldots,\uv_m$,
  \[
  \mathrm{Span}\ \{\uv_1,\uv_2,\ldots,\uv_m\}
  \]
  is a vector space.

  \pause
  \vspace{0.2in}
  Let's check if properties V1, V2, and V3 are satisfied.
  \vspace{1.5in}
\end{frame}

\begin{frame}
  \frametitle{Solutions to homogeneous linear equations is a vector space}
  Consider a set ${\mathcal S}$ of all $n$-vectors in the form $[x_1,x_2,\ldots,x_n]$ where
  \[
  \begin{array}{rcl}
    a_{11}\cdot x_1 + a_{12}\cdot x_2 + \cdots + a_{1n}\cdot x_n &=& 0 \\
    a_{21}\cdot x_1 + a_{22}\cdot x_2 + \cdots + a_{2n}\cdot x_n &=& 0 \\
    \vdots &=& \vdots \\
    a_{m1}x\cdot _1 + a_{m2}\cdot x_2 + \cdots + a_{mn}\cdot x_n &=& 0
  \end{array}
  \]
  \vspace{0.2in}
  Let's check if properties V1, V2, and V3 are satisfied.
  \vspace{1.5in}
  
\end{frame}

\begin{frame}
  \frametitle{Dot product}

  \begin{block}{Definition}
    For $n$-vectors $\uv=[u_1,u_2,\ldots,u_n]$ and $\vv=[v_1,v_2,\ldots,v_n]$, the {\bf dot product} of $\uv$ and $\vv$, denoted by $\uv\cdot\vv$, is
    \[
    u_1\cdot v_1 + 
    u_2\cdot v_2 +
    \cdots +
    u_n\cdot v_n 
    \]
  \end{block}

  \vspace{0.2in}
  \pause
  Using dot products, the previous set $\mathcal S$ can be written as
  \[
  \{ \vect{x}\in \rf^n :
  \vect{a}_1\cdot \vect{x} = 0,
  \vect{a}_2\cdot \vect{x} = 0,\ldots,
  \vect{a}_m\cdot \vect{x} = 0
  \}
  \]
  and we know that $\mathcal S$ is a vector space.
\end{frame}

\begin{frame}
  \frametitle{Parity-check code}

  \pause

  From message $\vect{a}=[a_1,a_2,a_3,a_4]$, we compute (in $GF(2)$)
  the parity check bit
  \[
  b = a_1 + a_2 + a_3 + a_4.
  \]
  \pause
  Now our encoded message becomes
  \[
  [a_1,a_2,a_3,a_4,a_5],
  \]
  where $a_5=b=a_1 + a_2 + a_3 + a_4$.  It can detects a single-bit
  error.
  \pause

  \vspace{0.2in}

  What can we say about the condition on $a_5$? \pause

  It is in fact a homogeneous linear equation (in $GF(2)$):
  \[
  a_1 + a_2 + a_3 + a_4+a_5 = 0
  \]
  \pause
  \vspace{0.1in}
  Now, what is the set of all possible codewords?
\end{frame}

\begin{frame}
  \frametitle{Hamming code}

  You can detect and correct more errors with Hamming codes.  In this
  version called a $[7,4]$ Hamming code, you encode $4$-bit data
  $[a_1,a_2,a_3,a_4]$ into a $7$-bit codeword
  $[a_1,a_2,a_3,a_4,a_5,a_6,a_7]$.  Using the formula:
  \[
  \begin{array}{rcl}
    a_5 &=& a_1 + a_2 + a_4\\
    a_6 &=& a_1 + a_3 + a_4\\
    a_7 &=& a_2 + a_3 + a_4
  \end{array}
  \]
  \pause

  Let's see how this works.

  \vspace{1.5in}
  
\end{frame}

\begin{frame}
  \frametitle{Parity check}
  Let
  \[
  \begin{array}{rcl}
    s_1 &=& a_1 + a_2 + a_4 + a_5\\
    s_2 &=& a_1 + a_3 + a_4 + a_6\\
    s_3 &=& a_2 + a_3 + a_4 + a_7
  \end{array}
  \]

  Given a codewords $\vect{w}=[c_1,c_2,\ldots,c_7]$, if we compute
  $s_1,s_2,s_3$, we would get all zero's.

  \pause

  What if there is an error?  Let's try.
  \vspace{1.5in}
\end{frame}

\begin{frame}
  \frametitle{Codewords from Hamming code}

  Turning the formula for $a_5,a_6,a_7$ around, we have $3$
  homogeneous linear equations:
  \[
  \begin{array}{rcl}
    a_1 + a_2 + a_4 + a_5 &=& 0\\
    a_1 + a_3 + a_4 + a_6 &=& 0\\
    a_2 + a_3 + a_4 + a_7 &=& 0
  \end{array}
  \]
  and again the set of all possible codewords $\W$ forms a vector
  space over $GF(2)$.
  
\end{frame}

\begin{frame}
  \frametitle{An object not passing through the origin: 2 dimensions}
\end{frame}

\begin{frame}
  \frametitle{An object not passing through the origin: 3 dimensions}
\end{frame}

\begin{frame}
  \frametitle{Translation}

  If we have a line or a plane passing through a vector $\vect{a}$,
  but not through the origin, how can we represent it?

  \pause

  \begin{itemize}
  \item Translate the object so that it passes through the origin.
    \pause
  \item We obtain a vector space $\V$.
    \pause
  \item Then we translate it back so that it passes through $\vect{a}$.
    \pause
  \item We get the set
    \[
    {\mathcal A} = \{ \vect{a} + \vect{u} : \vect{u} \in \V \}
    \]
    \pause
  \item {\em Question:} Is $\mathcal A$ a vector space?
    \pause
  \item We also write it as $\vect{a} + \V$.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Affine spaces}

  \begin{block}{Definition}
    If $\vect{a}$ is a vector and $\V$ is a vector space, then
    \[
    \vect{a} + \V
    \]
    is an \textcolor{red}{\bf affine space}.
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{An affine space and convex combination: 2 dimensions}
\end{frame}

\begin{frame}
  \frametitle{An affine space and convex combination: 3 dimensions}
\end{frame}

\begin{frame}
  \frametitle{Affine combination}
  
  \begin{block}{Definition}
    For any scalars $\alpha_1,\alpha_2,\ldots,\alpha_m$
    such that 
    \[
    \alpha_1 + \alpha_2 + \ldots + \alpha_m = 1
    \]
    and vectors $\uv_1,\uv_2,\ldots,\uv_m$, we say that a linear combination
    \[
    \alpha_1\uv_1 + \alpha_2\uv_2 + \cdots + \alpha_m \uv_m
    \]
    is an \textcolor{red}{\bf affine combination} of $\uv_1,\ldots,\uv_m$.
  \end{block}

  \pause

  \begin{block}{Definition}
    The set of all affine combinations of 
    vectors $\uv_1,\uv_2,\ldots,\uv_m$
    is called the \textcolor{red}{\bf affine hull} of
    $\uv_1,\uv_2,\ldots,\uv_m$.
  \end{block}

\end{frame}

\begin{frame}
  \frametitle{Convex combination: review}
  
  \begin{block}{Definition}
    For any scalars $\alpha_1,\alpha_2,\ldots,\alpha_m \geq 0$
    such that 
    \[
    \alpha_1 + \alpha_2 + \ldots + \alpha_m = 1
    \]
    and vectors $\uv_1,\uv_2,\ldots,\uv_m$, we say that a linear combination
    \[
    \alpha_1\uv_1 + \alpha_2\uv_2 + \cdots + \alpha_m \uv_m
    \]
    is a \textcolor{red}{\bf convex combination} of $\uv_1,\ldots,\uv_m$.
  \end{block}

  \begin{block}{Definition}
    The set of all convex combinations of 
    vectors $\uv_1,\uv_2,\ldots,\uv_m$
    is called the \textcolor{red}{\bf convex hull} of
    $\uv_1,\uv_2,\ldots,\uv_m$.
  \end{block}

\end{frame}

\begin{frame}
  \frametitle{Writing an affine space using a span}
  \pause

  \begin{block}{An affine space}
    An affine space passing through $\uv_1,\uv_2,\ldots,\uv_n$ is
    \[
    \uv_1 + \mathrm{Span}\ \{\uv_2-\uv_1,\uv_3-\uv_1,\ldots,\uv_n-\uv_1\}.
    \]
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Non-homogeneous linear system}
  Two linear systems:
  \[
  \begin{array}{rcl}
    \vect{a_1}\cdot\vect{x} &=& b_1 \\
    \vect{a_2}\cdot\vect{x} &=& b_2 \\
    &\vdots&\\
    \vect{a_m}\cdot\vect{x} &=& b_m
  \end{array}
  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
  \begin{array}{rcl}
    \vect{a_1}\cdot\vect{x} &=& 0 \\
    \vect{a_2}\cdot\vect{x} &=& 0 \\
    &\vdots&\\
    \vect{a_m}\cdot\vect{x} &=& 0
  \end{array}
  \]

  What can you say about the solution sets of these two related linear
  systems?

  \pause

  $\vect{0}$ is always a solution to the linear system on the right.

  
  Note: A linear equation whose right-hand-side is zero is called a
  {\bf homogeneous linear equation}.  A system of linear homogeneous
  equations is called a {\bf homogeneous linear system}.
\end{frame}

\begin{frame}
  \frametitle{Solutions of the two systems}

  Recall that if $\uv_1$ and $\uv_2$ are both solutions to the
  non-homogeneous linear system, we have that for any $i$
  \[
  \vect{a}_i \uv_1 - \vect{a}_i \uv_2
  = b_i - b_i = 0 = \vect{a}_i (\uv_1 - \uv_2).
  \]
  \pause

  This implies that $\uv_1-\uv_2$ is a solution to the homogeneous
  linear system.
\end{frame}

\begin{frame}
  Suppose that $\W$ is the set of all solution to the non-homogeneous
  linear system, i.e.,
  \[
  \W = \{\vect{x} : \vect{a}_i\vect{x} = b_i, \ \mbox{for $1\leq i\leq m$}\},
  \]
  and let $\uv\in \W$ be one of the solutions, we have that
  \[
  \{\vv - \uv : \vv\in\W \} 
  \]
  \pause
  is a vector space, because
  \pause
  \[
  \{\vv - \uv : \vv\in\W \}
  =
  \{\vect{x} : \vect{a}_i\vect{x} = 0, \ \mbox{for $1\leq i\leq m$}\}
  \]

  \vspace{0.1in}
  \pause

  In other words,
  \[
  \begin{array}{rcl}
    \W &=& \uv + \{\vv - \uv : \vv\in\W \}  \\
    &=& \uv + \{\vect{x} : \vect{a}_i\vect{x} = 0, \ \mbox{for $1\leq i\leq m$}\},
  \end{array}
  \]
  \pause
  i.e., $\W$ is an affine space.
\end{frame}

\begin{frame}
  \frametitle{Solutions to a non-homogeneous linear system}

  \begin{lemma}
    If the solution set of a linear system is not empty, it is an
    affine space.
  \end{lemma}
\end{frame}
